Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM and SVA normlaisation)

### 2016-09-22 code/Rmd/CMC_DLPFC_StarAligned_TMM_SVA.Rmd
Modified DLPFC with SVA pipeline to incorporate custom desing

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
                  "DLPFC_RNA_report: Library Batch")

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Institution", "Gender", "Ethnicity", "CDR", "CauseOfDeath", "IsolationBatch", "LibraryBatch", "Dx")
ContCovariates <- c("Age_of_Death", "PMI", "pH", "RIN")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'IsolationBatch')

@@

# Residual covariates to choose from
residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES[, residCovars, drop=F], 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch))

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)

writeLines('Following samples are missing PMI information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(PMI))

writeLines('Following samples are missing gender information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(Gender))

writeLines('Following samples are missing RIN information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(RIN))

writeLines('Following samples are tagged as outliers')
METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 
                                                   'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                                                   'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 
                                                   'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Institution", "Gender", "LibraryBatch", "Dx")
ContCovariates <- c("Age_of_Death", "PMI", "RIN")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

### Normlaisation (Custom design)
To be consistent with DLPFC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender
```{r custom.design}
# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()

@@

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
